---
title: 'STAT 110 Section 03: Conditioning'
author: "Al Xin"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Materials are availabile in Section Materials > Al on Canvas or on GitHub at https://www.github.com/awqx/stat110-2021. 

## Overview

Topics covered this week: 

- Conditional probability
- Bayes' rule
- Law of total probability
- Independence
- Conditional independence

## Conditional probability

R has several available datasets, viewable using `data()`. For this section, we will be using `iris` and `HairEyeColor`.

### Iris

```{r}
# Run the following to view the `iris` dataset
iris
```

First, let's examine the dataset. If you use the function `View(iris)`, the resulting table has more convenient sorting properties than looking through the dataset in the console. 

```{r}
View(iris)
```

We can also create a visual representation of the data. Choose any two quantitative variables and create a plot, coloring each dot according to the iris species. 

```{r}
library(ggplot2)
```

We should be able to see some clustering of the variables. This clustering lends itself to some conditioning. 

Let's compare the probability that a flower is a member of any species given its physical characteristics. 

First, investigate the two columns of `Petal.Width` and `Species`. Do you notice any trends? 

```{r}
library(dplyr)
View(select(iris, Petal.Width, Species))

# There's a more "intuitive" way to run the above command
```

Can you think of a conditional probability rule that maximizes the probability that a flower is a `setosa` given a condition?

What about `versicolor` and `virginica`?

### Hair and Eye Color

Like with `iris`, view the data to get an idea of what it looks like. Have we seen this type of dataset before? Experiment with indexing this type of data. 

```{r}
HairEyeColor
View(HairEyeColor)
```

Try casting this `table` to a `data.frame` for easier sorting with our functions. Alternatively, if you know the labels for the rows and columns, you can complete the conditional probabilities using that method. 

What is the conditional probability that a student has green eyes given that they have black hair?

What is the conditional probability that a student with blue eyes is female?

## Law of total probability

Let's look at `HairEyeColor` again.

- What is the unconditional probability that a student has black hair?
- What is the unconditional probability that a student is female?
- What is the unconditional probability that a student is male?

Check if these approaches are equivalent:

- What is the sum of the intersection of probabilities of (black hair, female) and (black hair, male)?
- What is the sum of the products of the conditional probabilities (LOTP)?

## Conditional independence

Take a look at the data `Titanic`. What does this data represent?

Answer the following questions:

- Is survival independent of passenger class?
- Is survival independent of age?

## Applied R

### Choosing cards

In R, we can compute the binomial coefficient with the expression `choose`. For example, if determining the number of ways to select 5 cards from a pack of 52, we would run the following code: 

```{r}
choose(52, 5)
```

R also has the function `factorial`. Try calculating the binomial coefficient with all terms included and make sure that it agrees with your previous answer.

```{r}
factorial(1)
```

Let's apply this to a particular function to a card-choosing problem. What is the probability of selecting a full house in a random selection of 5 cards from a 52 card deck?

```{r}
choose()
```

Functions like choose can also be vectorized. This may be helpful in checking one of the homework problems. See an example below:

```{r}
choose()
```

